Modal Abstractions for Virtualizing Memory Addresses

### 2023-04-05 programlogic.tex
+

\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
\subsubsection{Issues with Using Hoare Triples in Address-Space Specification}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3\mapsto_{r} \;r_1 \ast r \mapsto_{r} \;r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3 \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3 \mapsto_{r} r_1 \ast r\mapsto_{r} r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3\mapsto_{r}r_1 \ast r \mapsto_{r}r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.

@@

Readers familiar with dynamic frames~\cite{parkinson2011relationship} might find it useful to notice that a different perspective on this matter is that virtual points-to assertions are self-stable \emph{except} for changes in \lstinline|cr3|, so framing would then naturally require other means of holding \lstinline|cr3| constant (or saving and restoring it).
Virtual points-to assertions could be made self-stable by also giving them partial ownership over \lstinline|cr3| assertions, but this would require explicitly plumbing that ownership from \emph{all} assertions back to any place an address space change might occur; this would seem to be a far graver loss of modularity than this extra quirk in framing discussions.
\subsubsection{Defining Hoare Doubles for Address-Space Specification}

\begin{figure}
  \[
  \begin{array}{l}
    \textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rootaddr } : \textsf{iProp }\Sigma := \\
   \qquad ((\textsf{cr cr3}) \mapsto_{\textsf{r}} (\textsf{num rootaddr)}) \textsf{rootaddr}) \ast (\Phi \textsf{ rootaddr}) \wand \textsf{WP e E } @ \textsf{s ; E1} \{\{\_, \textsf{True} \}\}
    \end{array}
  \]
\caption{Unfolded Hoare-Double Definition for \textsf{vProp} Logic }
\label{fig:wpddefinition}
\end{figure}

\begin{figure}
\begin{mathpar}
\inferrule[WriteToRegFromReg]{
  \{P \ast r_d \mapsto_{r} \textsf{rvs} \ast r_s \mapsto_{r}\{q\} \textsf{ rvs} \}\;\overline{ is}
}{
  \{P \ast r_d \mapsto_{r} \textsf{rvd} \ast r_s \mapsto_{r}\{q\} \textsf{ rvs} \}
  \textsf{ mov}~\textsf{r}_d,~\textsf{r}_s;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\\
\inferrule[WriteToMemFromReg]{
  \{P \ast r_s \mapsto_{r}\{q\}  \textsf{rvs} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \ast r_a \mapsto_{r} \{q\} \textsf{ vaddr} \ast (\textsf{vaddr} \mapsto_{\textsf{t}} \textsf{v})\;\textsf{cr3val} \}\;\overline{ is}  
}{
  \{P \ast r_s \mapsto_{r}\{q\}  \textsf{rvs} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \ast r_a \mapsto_{r}\{q\} \textsf{ vaddr} \ast (\textsf{vaddr} \mapsto_{\textsf{t}} \textsf{rvs})\;\textsf{cr3val} \}
  \textsf{ mov}~\textsf{r}_a,~\textsf{r}_s;\;\overline{is}
}
\\
\inferrule[WriteToRegFromMem]{
  \{P \ast r_d \mapsto_{r}  \textsf{v} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \ast r_a \mapsto_{r} \{q\} \textsf{ vaddr} \ast (\textsf{vaddr} \mapsto_{\textsf{t}} \textsf{v})\;\textsf{cr3val} \}\;\overline{is}
}{
  \{P \ast r_d \mapsto_{r}  \textsf{rvd} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \ast r_a \mapsto_{r} \{q\} \textsf{ vaddr} \ast (\textsf{vaddr} \mapsto_{\textsf{t}} \textsf{v})\;\textsf{cr3val} \}
  \textsf{mov}~\textsf{r}_d,~\textsf{r}_a;\;\overline{is}
}
\\
\inferrule[WriteToRegCtlFromReg]{
  \{P \ast r_s \mapsto_{r}\{q\}  \textsf{ rvs} \ast \textsf{cr3} \mapsto_{r} \textsf{rvs}  \}\overline{is}
}{
  \{P \ast r_s \mapsto_{r}\{q\}  \textsf{ rvs} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \}
  \textsf{mov}~\textsf{cr3},~r_s;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
\caption{Selected Reasoning Rules for ..}
\label{fig:wpdamd}
\end{figure}
\paragraph{Register To Memory}
\paragraph{Memory To Register}
\paragraph{Register To Register}
\todo[inline,color=red]{STOP HERE}

>>>>

\subsection{Considering Hoare Doubles in the Context of Switching Address-Space}
\subsubsection{Issues with Using Hoare Triples in Address-Space Specification}
\label{sec:issues}
An important subtlety arises with supporting \lstinline|mov|s into \lstinline|%cr3|. Consider the hypothetical rule:
\begin{mathpar}
\inferrule[Broken]{ }{
  \{P \ast cr3\mapsto_{r} \;r_1 \ast r \mapsto_{r} \;r_2 \ast [r_2](Q)\}
  \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
  \{[r_1](P) \ast cr3 \mapsto_{r} r_2 \ast r \mapsto_{r} r_2 \ast Q\}
}
\end{mathpar}
This rule captures the intuitive change of address space in a Hoare triple, rather than double, form. The problem with this is that it interacts quite poorly with the traditional frame rule and the modal flavor of virtual points-to assertions:
\begin{mathpar}
  \inferrule*[right=Frame]{
    \inferrule*[right=Broken]{ }{
    \{\mathsf{emp} \ast cr3 \mapsto_{r} r_1 \ast r\mapsto_{r} r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{[r_1](\mathsf{emp}) \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast Q\}
    }
  }{
    \{a\mapsto_\mathsf{v} x \ast \mathsf{emp} \ast cr3\mapsto_{r}r_1 \ast r \mapsto_{r}r_2 \ast [r_2](Q)\}
    \texttt{mov}~\texttt{\%cr3},~r%\lstinline|mov %cr3, r| 
    \{a\mapsto_\mathsf{v} x \ast [r_1](\mathsf{emp}) \ast cr3 \mapsto_{r} r_1 \ast r \mapsto_{r} r_2 \ast Q\}
  }
\end{mathpar}
Notice that both the precondition and postcondition assert that $a\mapsto_\mathsf{v} x$ in the current address space, but we have no basis for concluding that address translation is preserved by the change of address space. So this derivation clearly leads to an unsound conclusion. This suggestss that the traditional frame rule and the Hoare triple presentation of the change-of-address-space rule cannot soundly coexist in the same system.
The heart of the problem is that while updating \lstinline|cr3| is \emph{physically} local, it globally changes the interpretation of virtual addresses. So it is simply unsound to frame around \lstinline|cr3| updates.

@@

Readers familiar with dynamic frames~\cite{parkinson2011relationship} might find it useful to notice that a different perspective on this matter is that virtual points-to assertions are self-stable \emph{except} for changes in \lstinline|cr3|, so framing would then naturally require other means of holding \lstinline|cr3| constant (or saving and restoring it).
Virtual points-to assertions could be made self-stable by also giving them partial ownership over \lstinline|cr3| assertions, but this would require explicitly plumbing that ownership from \emph{all} assertions back to any place an address space change might occur; this would seem to be a far graver loss of modularity than this extra quirk in framing discussions.
\subsubsection{Defining Hoare Doubles for Address-Space Specification}
Based on the issues we discuss in Section \ref{sec:issues}, and the design of the abstractions for \textsf{vProp} logic, e.g. virtual-pointsto assertion and the modal address-space abstraction, we give the unfolded Hoare-double definition in Figure \ref{fig:wpddefinition}. 
\begin{figure}
  \[
  \begin{array}{l}
    \textsf{wpd\_def e s E1 } \Phi \;\mathsf{ rootaddr } : \textsf{iProp }\Sigma := \\
   \qquad ((\textsf{cr cr3}) \mapsto_{\textsf{r}} (\textsf{num rootaddr)}) \textsf{rootaddr}) \ast (\Phi \textsf{ rootaddr}) \wand \textsf{WP e E } @ \textsf{s ; E1} \{\{\_, \textsf{True} \}\}
    \end{array}
  \]
\caption{Unfolded Hoare-Double Definition for \textsf{vProp} Logic }
\label{fig:wpddefinition}
\end{figure}
This definition carries only the facts ($\Phi$) that are valid under the current addresss-space -- $\Phi \; \textsf{rootaddr}$. It also helps to the agnosticism of the client on what the current address-space is by carrying the \textsf{cr3} value on the definition. 
\begin{figure}
\begin{mathpar}
\inferrule[WriteToRegFromReg]{
  \{P \ast r_d \mapsto_{r} \textsf{rvs} \ast r_s \mapsto_{r}\{q\} \textsf{ rvs} \}\;\overline{ is}
}{
  \{P \ast r_d \mapsto_{r} \textsf{rvd} \ast r_s \mapsto_{r}\{q\} \textsf{ rvs} \}
  \textsf{ mov}~\textsf{r}_d,~\textsf{r}_s;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\\
\inferrule[WriteToMemFromReg]{
  \{P \ast r_s \mapsto_{r}\{q\}  \textsf{rvs} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \ast r_a \mapsto_{r} \{q\} \textsf{ vaddr} \ast (\textsf{vaddr} \mapsto_{\textsf{t}} \textsf{v})\;\textsf{cr3val} \}\;\overline{ is}  
}{
  \{P \ast r_s \mapsto_{r}\{q\}  \textsf{rvs} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \ast r_a \mapsto_{r}\{q\} \textsf{ vaddr} \ast (\textsf{vaddr} \mapsto_{\textsf{t}} \textsf{rvs})\;\textsf{cr3val} \}
  \textsf{ mov}~\textsf{r}_a,~\textsf{r}_s;\;\overline{is}
}
\\
\inferrule[WriteToRegFromMem]{
  \{P \ast r_d \mapsto_{r}  \textsf{v} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \ast r_a \mapsto_{r} \{q\} \textsf{ vaddr} \ast (\textsf{vaddr} \mapsto_{\textsf{t}} \textsf{v})\;\textsf{cr3val} \}\;\overline{is}
}{
  \{P \ast r_d \mapsto_{r}  \textsf{rvd} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \ast r_a \mapsto_{r} \{q\} \textsf{ vaddr} \ast (\textsf{vaddr} \mapsto_{\textsf{t}} \textsf{v})\;\textsf{cr3val} \}
  \textsf{mov}~\textsf{r}_d,~\textsf{r}_a;\;\overline{is}
}
\\
\inferrule[WriteToRegCtlFromReg]{
  \{P \ast r_s \mapsto_{r}\{q\}  \textsf{ rvs} \ast \textsf{cr3} \mapsto_{r} \textsf{rvs}  \}\overline{is}
}{
  \{P \ast r_s \mapsto_{r}\{q\}  \textsf{ rvs} \ast \textsf{cr3} \mapsto_{r} \textsf{cr3val}  \}
  \textsf{mov}~\textsf{cr3},~r_s;\;\overline{is}
  %\lstinline|mov %cr3, r| 
}
\end{mathpar}
\caption{Selected Reasoning Rules for ..}
\label{fig:wpdamd}
\end{figure}
\paragraph{Register To Memory}
\paragraph{Memory To Register}
\paragraph{Register To Register}
\todo[inline,color=red]{STOP HERE}
